A tutorial of Bayesian beta regressions with brms in R

### 2025-02-04 manuscript/manuscript.qmd
edit case study 1

---
title: A tutorial of Bayesian beta regressions with brms in R
author:
  - name: Stefano Coretta
    email: s.coretta@ed.ac.uk
    affiliations: 
        - id: uoe
          name: University of Edinburgh
          department: Linguistics and English Language
          address: 3 George Sq
          city: Edinburgh
          state: United Kingdom
          postal-code: EH8 9AD
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Paul Bürkner
    email: paul.buerkner@gmail.com
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: 
  - linguistics.bib
  - statistics.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

@@

Here is what the tibble looks like.

```{r}
#| label: ita-egg

ita_egg
```

@@

The `summary()` function prints the full model summary. For conciseness, we will use the `fixef()` function which prints the regression coefficients. \[EXPLAIN EXPECTED PREDICTIONS XXX\]. The full summary with an explanation of each part can be found in XXX. For each coefficient in the model, `fixef()` prints the name of the coefficient, the mean estimate, the estimate error and the lower and upper limits of a Bayesian Credible interval (CrI). Here, we print an 80% CrI. There is nothing special about 95% CrI within Bayesian inference and instead experts recommend to check and report a variety of CrIs. Bayesian CrIs indicate that at a certain probability levels the "true" estimate lies within that interval: so, for example, a 90% CrI \[A, B\] indicates that there is a 90% probability that the "true" estimate is between A and B. Different probability levels correspond to different levels of confidence: the higher the probability the higher the confidence (always conditional on data and model). Obtaining CrIs at different probability levels allows researchers to make more fine-grained inferential statements than the frequentist significance dichotomy affords. For simplicity of exposition, we will use 80% CrIs in this case study but we strongly recommend researchers to always obtain CrIs at different levels of probability and base their inferences on all and not one in particular. To reiterate, in Bayesian inference, an 80% CrI indicates the range of values within which the true estimate falls at 80% probability or confidence. We round all values to the nearest 2 digits for clarity.

```{r}
#| label: voi-prop-bm-fixef
round(
  fixef(voi_prop_bm, prob = c(0.1, 0.9)),
  digits = 2
)
```

@@

Before being able to plot the predictions, it's important to get familiar with the so-called posterior draws. \[Bayesian MCMC XXX\]. Posterior draws can be conveniently obtained with the `as_draws_df()`. For the moment, we will extract only the draws of the constant regression coefficients (model variables starting with `b_`). To check which coefficients are available in a model, use `get_variables()` from the tidybayes package. `as_draws_df()` returns a tibble where each column contains the drawn values of a coefficient. The probability distribution of the drawn values of each coefficient is the posterior probability distribution of that coefficient. Note that, due to the indexing coding of vowel and C2, all coefficient except `b_speech_rate_c` are *predicted log-odds* for each vowel/C2 combination (the drawn values for `b_speech_rate_c` are drawn *differences* in log-odds for each unit increase of speech rate). The drawn values are in log-odds, but we can convert them to proportions with `plogis()` (we will do this when plotting below).

@@

# extract only coefficient variables starting with "b_"
voi_prop_bm_draws <- as_draws_df(voi_prop_bm, variable = "^b_", regex = TRUE)
voi_prop_bm_draws
```

We can now wrangle this tibble and plot the posterior distributions for each vowel/C2 combination.

```{r}
#| label: voi-prop-bm-draws-long

voi_prop_bm_draws_long <- voi_prop_bm_draws |> 
  # drop b_speech_rate_c before pivoting
  select(-b_speech_rate_c) |> 
  # pivot vowel:c2 columns
  pivot_longer(`b_voweli:c2k`:`b_vowelu:c2p`, names_to = "coeff") |> 
  # separate "coeff" labels into type ("b"), vowel and c2
  separate(coeff, into = c("type", "vowel", "c2"))

voi_prop_bm_draws_long
```

For plotting, we can use ggplot2 statistics layers from the ggdist package. `stat_halfeye()` plots the density of the posterior probability (in grey), its median (point) and CrIs (lines). Let's use 60 and 80% CrIs and transform the log-odds values to proportions with `plogis()`.

```{r}
#| label: fig-voi-prop-bm-1

@@

voi_prop_bm_draws_long_c2 <- voi_prop_bm_draws_long |> 
  # grouping by .draw and c2 ensures that averaging applies only within draw and c2
  group_by(.draw, c2) |> 
  # calculate the mean value within draw/c2
  summarise(
    value_mean = mean(value), .groups = "drop"
  )

voi_prop_bm_draws_long_c2 |> 
  ggplot(aes(plogis(value_mean), c2)) +
  stat_halfeye(.width = c(0.6, 0.8))
```

Based on the posterior distributions of the mean voicing proportion by consonant, /p/ has a somewhat higher voicing proportion than /k/ and /t/. The real question is: how much higher? We can quantify this by taking the difference of the drawn values for /p/ and those for /t, k/. Since we want to compare /t, k/ to /p/, we should first average the draws of /t, k/ and then take the difference of the averaged draws and the draws of /p/.

@@

voi_prop_bm_diff <- voi_prop_bm_draws_long_c2 |> 
  # pivot data to create one column per consonant with the mean drawn values,
  # with one draw per raw
  pivot_wider(names_from = c2, values_from = value_mean) |> 
  mutate(
    # calculate the mean of /k/ and /t/, for each draw
    c2tk = mean(c(c2k, c2t)),
    # calculate the difference of /p/ and /t, k/
    c2p_tk_diff = c2p - c2tk
  )

voi_prop_bm_diff
```

```{r}
#| label: fig-diff-p-tk

>>>>

---
title: A tutorial of Bayesian beta regressions with brms in R
author:
  - name: Stefano Coretta
    email: s.coretta@ed.ac.uk
    affiliations: 
        - id: uoe
          name: University of Edinburgh
          department: Linguistics and English Language
          address: 3 George Sq
          city: Edinburgh
          state: United Kingdom
          postal-code: EH8 9AD
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Paul Bürkner
    email: paul.buerkner@gmail.com
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: 
  - linguistics.bib
  - statistics.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Journal Name
      formatting: preprint
      model: 3p
      cite-style: authoryear
execute:
  cache: true
---

@@

@tbl-ita-egg shows the first ten rows of the tibble (only relevant columns are included).

```{r}
#| label: tbl-ita-egg
#| tbl-cap: "First 10 rows of the data of voicing proportion within stop closure."
#| echo: false

ita_egg |> select(speaker, word, vowel, c2, voi_clo_prop) |> head(n = 10) |> knitr::kable()
```

@@

The `summary()` function prints the full model summary. For conciseness, we will use the `fixef()` function which prints the regression coefficients. \[EXPLAIN EXPECTED PREDICTIONS XXX\]. The full summary with an explanation of each part can be found in XXX. @tbl-voi-prop-fixef reports the output of `fixef()` as a table (we round all values to the nearest 2 digits for clarity.). For each coefficient in the model, `fixef()` prints the name of the coefficient, the mean estimate, the estimate error and the lower and upper limits of a Bayesian Credible interval (CrI). Here, we print an 80% CrI. There is nothing special about 95% CrI within Bayesian inference and instead experts recommend to check and report a variety of CrIs. Bayesian CrIs indicate that at a certain probability levels the "true" estimate lies within that interval: so, for example, a 90% CrI \[A, B\] indicates that there is a 90% probability that the "true" estimate is between A and B. Different probability levels correspond to different levels of confidence: the higher the probability the higher the confidence (always conditional on data and model). Obtaining CrIs at different probability levels allows researchers to make more fine-grained inferential statements than the frequentist significance dichotomy affords. For simplicity of exposition, we will use 80% CrIs in this case study but we strongly recommend researchers to always obtain CrIs at different levels of probability and base their inferences on all and not one in particular. To reiterate, in Bayesian inference, an 80% CrI indicates the range of values within which the true estimate falls at 80% probability or confidence.

```{r}
#| label: voi-prop-fixef
#| tbl-cap: "Regression coefficients of a beta regression of voicing proportion (`voi_prop_bm`)."
#| eval: false

fixef(voi_prop_bm, prob = c(0.1, 0.9))
```

```{r}
#| label: tbl-voi-prop-fixef
#| tbl-cap: "Regression coefficients of a beta regression of voicing proportion (`voi_prop_bm`)."
#| echo: false

round(
  fixef(voi_prop_bm, prob = c(0.1, 0.9)),
  digits = 2
) |> knitr::kable()
```

@@

Before being able to plot the predictions, it's important to get familiar with the so-called posterior draws. \[Bayesian MCMC XXX\]. Posterior draws can be conveniently obtained with the `as_draws_df()`. For the moment, we will extract only the draws of the constant regression coefficients (model variables starting with `b_`). To check which coefficients are available in a model, use `get_variables()` from the tidybayes package. `as_draws_df()` returns a tibble where each column contains the drawn values of a coefficient. @tbl-voi-prop-draws shows the first ten rows and first five columns of the output of `as_draws_df()`. The probability distribution of the drawn values of each coefficient is the posterior probability distribution of that coefficient. Note that, due to the indexing coding of vowel and C2, all coefficient except `b_speech_rate_c` are *predicted log-odds* for each vowel/C2 combination (the drawn values for `b_speech_rate_c` are drawn *differences* in log-odds for each unit increase of speech rate). The drawn values are in log-odds, but we can convert them to proportions with `plogis()` (we will do this when plotting below).

@@

# extract only coefficient variables starting with "b_"
voi_prop_bm_draws <- as_draws_df(voi_prop_bm, variable = "^b_", regex = TRUE)
```

```{r}
#| label: tbl-voi-prop-draws
#| tbl-cap: "First ten rows and 5 columns of the posterior draws for the model `voi_prop_bm`."
#| echo: false

voi_prop_bm_draws[,1:5] |> head(n = 10) |> knitr::kable()
```

We can now wrangle this tibble and plot the posterior distributions for each vowel/C2 combination. @tbl-voi-long shows the first ten rows.

```{r}
#| label: voi-prop-bm-draws-long
#| warning: false

voi_prop_bm_draws_long <- voi_prop_bm_draws |> 
  # drop b_speech_rate_c before pivoting
  select(-b_speech_rate_c) |> 
  # pivot vowel:c2 columns
  pivot_longer(`b_voweli:c2k`:`b_vowelu:c2p`, names_to = "coeff") |> 
  # separate "coeff" labels into type ("b"), vowel and c2
  separate(coeff, into = c("type", "vowel", "c2"))
```

```{r}
#| label: tbl-voi-long
#| tbl-cap: "First ten rows of posterior draws from `voi_prop_bm` in long format."
#| echo: false

voi_prop_bm_draws_long |> head(n = 10) |> knitr::kable()
```

For plotting, we can use ggplot2 statistics layers from the ggdist package. `stat_halfeye()` plots the density of the posterior probability (in grey), its median (point) and CrIs (lines). Let's use 60 and 80% CrIs and transform the log-odds values to proportions with `plogis()`. See @fig-voi-prop-bm-1.

```{r}
#| label: fig-voi-prop-bm-1
#| fig-cap: "Expected predictions of voicing proportion by vowel and consonant from `voi_prop_bm`."
#| message: false

@@

voi_prop_bm_draws_long_c2 <- voi_prop_bm_draws_long |> 
  # grouping by .draw and c2 ensures that averaging applies only within draw and c2
  group_by(.draw, c2) |> 
  # calculate the mean value within draw/c2
  summarise(
    value_mean = mean(value), .groups = "drop"
  )
```

```{r}
#| label: fig-voi-pbdl-c2
#| fig-cap: "Expected predictions of voicing proportion by consonant, averaged across vowels."
voi_prop_bm_draws_long_c2 |> 
  ggplot(aes(plogis(value_mean), c2)) +
  stat_halfeye(.width = c(0.6, 0.8))
```

Based on the expected posterior distributions of the mean voicing proportion by consonant, /p/ has a somewhat higher voicing proportion than /k/ and /t/. The real question is: how much higher? We can quantify this by taking the difference of the drawn values for /p/ and those for /t, k/ (all averaged across vowels). Since we want to compare /t, k/ with /p/, we should first average the average draws of /t, k/ and then take the difference of the averaged /t, k/ draws and the draws of /p/. @tbl-voi-pbd shows the first ten rows of the resulting data frame. The posterior distribution of the expected difference is shown in @fig-diff-p-tk.

@@

voi_prop_bm_diff <- voi_prop_bm_draws_long_c2 |> 
  # pivot data to create one column per consonant with the mean drawn values,
  # with one draw per raw
  pivot_wider(names_from = c2, values_from = value_mean) |> 
  mutate(
    # calculate the mean of /k/ and /t/, for each draw
    c2tk = mean(c(c2k, c2t)),
    # calculate the difference of /p/ and /t, k/
    c2p_tk_diff = c2p - c2tk
  )
```

```{r}
#| label: tbl-voi-pbd
#| tbl-cap: "First 10 rows of expected difference of voicing proportion between /t, k/ and /p/."
#| echo: false

voi_prop_bm_diff |> head(n = 10) |> knitr::kable()
```

```{r}
#| label: fig-diff-p-tk
#| fig-cap: "Posterior distribution of the difference between the expected voicing proportion of /t, k/ vs /p/, averaged across vowels."
